---
output:
  md_document:
    variant: gfm
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(knitr)
library(kableExtra)
library(tidyverse)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "90%",
  fig.width = 7*1.3,
  fig.height = 6*1.3,
  cache = T
)
theme_set(theme_bw())
```

# bpscm

## Overview

This repository contains the code to evaluate shared component models
(SCM) for bivariate count data which share common cases using a
Bivariate-Poisson distribution (BP-SCM). The method is described in
the paper entitled *Joint disease mapping for bivariate count data with
residual correlation due to unobserved common cases* submitted for publication in
Biostatistics.

## Content
The repository contains three folders:

* **stan** : `stan` code used to fit BP-SCM. 
  - `bp_scm_age.stan` is the main program ;
  - `join_fun.stan` regroups stan funcions used in `bp_scm_age.stan`,
   including BP likelihood and function for spatial analysis from
   https://github.com/ConnorDonegan/Stan-IAR/ . 
* **R** : `R` utility functions. 
  - `prep_data.R` prepares data for `stan` use;
  - `post.R` recovers and formats posterior estimates for key BP-SCM
    estimates. 
* **data** : data for demonstration. These data are used and
  described in detail in the simulation section of the paper
  - `lop.rds` : age-incidence rates for the $X_{ijk}$ derived using
  two proxies for lips-oral-pharynx (LOP) cancer incidence over the 2013-2019 period. 
  - `py.rds` : person-years for 611 zip-codes in east of France
    for the 2013-2019 period.
  - `shap_geo.rds` : shapefile for the 611 zip-codes, in `sf`
    format.
  - `sim_data.R` : `R` code used to generate `data_sim.rds`
  - `fit.rds` : results of BP-SCM fitted to `data_sim.rds` (see hereafter)
  - `data_sim.rds` : simulated bivariate counts for the 611
    zip-codes, using following parameters

<center>

|  |   N | $\delta_i$ | $\sigma_i$ | $\phi$ | 
|:-----|----:|-----:|-----:|----:|
| $X_1$   |   5 |  0.15 | 0.25 | 0.5 |
| $X_2$   |   7 |  0.10 | 0.15 | 0.7 |
| $X_3$   |  10 |  0.30 | | 0.8 |

</center>


## Illustration

As an illustration of how to use the proposed code, we will fit the
BP-SCM to the simulated data `data_sim.rds`.
We begin by loading packages and programs required for the analysis:

```{r}

library(rstan)
library(tidybayes)
library(tidyverse)
library(magrittr)
library(mgcv) 
library(INLA)
library(sf)
source("./R/prep_data.R")
source("./R/post.R")
theme_set(theme_bw())

``` 

### Prepare data for stan use

The `data_sim.rds` dataset is tabulated by geographical area, proxy
(Y1, Y2, X1, X2, X3) and age (columns `age` gives the median of the
5-years age class). `py` gives the number of person-years, `Y` the
observed counts, `lbd` the Poisson mean of the counts.

```{r}

dt <- readRDS("./data/data_sim.rds")
dt %>%
  filter(geo=="08000", age ==62.5)

``` 

`dt` contains the usually unobserved counts for the $X_{ijk}$. We
thus filter out those observations to focus on bivariate counts
$(Y_{1jk},Y_{2jk})$ to be fitted in the BP-SCM. 

```{r}

dt_Y <- dt %>%  filter(str_detect(prox,"Y")) 

``` 

To prepare the data for `stan` use, we have to define the different
component of the model (i.e., counts, proxy identifier, offsets, the type
of Gaussian Markov Random Field (GMRF) for the random effects $b_{ij}$
and $c_j$, spline basis for the age rates). This is done by
specifying a formulae for the model, as follow:

```{r}

form <-inc(Y,prox) ~ offset(log(py)) +  s(age,bs="cr",k=5) + 
  s(geo,X1,bs="bym2") +  s(geo,X2,bs="bym2") +  s(geo,X3,bs="bym2")

``` 

The term `inc(Y,prox)` indicates that the counts are in variable `Y`
and the proxy identifier is `prox`, the term `s(geo,X1,bs="bym")` specifies the
type of GMRF for the proxy $X_1$ (`bym2` here; alternatives are `iid`
or `icar`), and the spline is specified as in the `mgcv` package.

The formulae is used in `prep_data` function to prepare data needed to
evaluate the `stan` model. The geographical structure of the data is
specified thanks to the `geo` parameter of the function.

Note that the identifier of the geographical areas, in the shape file and in
the data set, must be named `geo`.

```{r}

shap_geo <- readRDS("./data/shap_geo.rds")
stan_dt <-  prep_data(dt = dt_Y,
            geo=shap_geo,
            form=form)

``` 

### Estimate the model 
The BP-SCM is then evaluated with usual `rstan` functions:

```{r, eval = F}

bpscm <- stan_model("./stan/bp_scm.stan", model_name = "bpscm")

chains <- 4
iter <- 2000
warmup <- 1000

deb<-Sys.time()
fit = 
  sampling(
    bpscm,
    data = stan_dt,
    chains = chains, cores = chains,
    iter = iter, warmup =warmup,
    pars = c("Int","rho","tau","w","theta","r","mu","nu","lbd","log_lik","sd_c","sd_b","phi","bc"),
    seed = 24,
    open_progress = FALSE,verbose=T
  )
(tps <- difftime(Sys.time(),deb,units = "mins"))

``` 


```{r, eval=T,echo=F}

fit <- readRDS("./data/fit.rds")

``` 


The model took about half an hour to run and indicated no pathological behaviour:

```{r}

check_hmc_diagnostics(fit)

``` 

### Posterior

The `get_pars` function retrieves posterior summary of main models
parameters. Parameters have no convergence issue ($\hat{R}<1.01$ and
ESS>400) and estimates are in line with the values used to simulate the data
(although with some variability):

```{r}

get_pars(fit,pars = c("sd_c","tau","sd_b","phi","w"))

``` 


The `get_rates`  function gives the posterior summary for age-rate
$\rho_{ijk}$ and $r_{ijk}$:

```{r}

post_rates <- get_rates(fit,stan_dt)

``` 

The estimated age-rates closely match the true age-rates used to
simulate the data:

```{r}

post_rates %>%
  ggplot()+
  aes(age, 10^5 * pm) +
  geom_line() +
  geom_ribbon( aes(ymin = 10^5 * low, ymax = 10^5 * up) ,
               alpha=0.2) +
  geom_line(data = dt %>% group_by(age,prox) %>% slice(1),
            aes(age,10^5 * ar),
            size=1,
            colour = "orange") +
  facet_wrap(~prox, scales="free") +
  ylab("Age-incidence rate (for 100 000 person)") +
  xlab("Age")


``` 

Last, we may look at the estimated relatives risks $\nu_{ij}$ and
$\theta_{ij}$. They can be extracted with the `get_rr` function:

```{r}

post_rr <- get_rr(fit,stan_dt)

``` 

These relatives risks may be compared with the true relatives risks
used to simulate the data:

```{r}

post_rr %<>%
  left_join(
      dt %>% 
        as_tibble() %>%
        group_by(geo,prox) %>% 
        summarise(tm=sum(lbd)/sum(py*ar), py = sum(py))
  ) 

post_rr %>%
  ggplot() +
  aes(tm,pm,size = py) +
  geom_point(alpha=.3) +
  geom_abline(intercept=0, slope = 1) +
  facet_wrap(~prox) +
  theme(legend.position = "bottom") +
  xlab("True value") + 
  ylab("Predicted value")

``` 

The shrinkage of the predicted risk for $X_1$ (with low number of
cases and low dependence on $c_j$) is important, whereas
the other risks looks well predicted by the model. 

This can be seen on maps of the predicted risks:

```{r, fig.width = 7.5, fig.height = 14}

lab_risk <- .%>% 
  mutate(prox =case_when(
    prox=="X1" ~"nu['1j']",
    prox=="X2" ~"nu['2j']",
    prox=="X3" ~"nu['3j']",
    prox=="Y1" ~"theta['1j']",
    prox=="Y2" ~"theta['2j']"
    ))

post_rr %>%
  left_join(shap_geo) %>%
  st_as_sf() %>%
  rename(True = tm, Predicted = pm) %>%
  gather(typ,rr, Predicted, True) %>%
  lab_risk %>%
  ggplot( data = ., aes(fill=rr)) +
  geom_sf( colour = NA) +
  scale_fill_gradientn(colours=rev(RColorBrewer::brewer.pal( n = 9, name = "RdYlGn")),
                       na.value = "transparent",
                       limits=c(0.7,1.3),
                       oob = scales::squish)+
  facet_grid(prox~typ, labeller = label_parsed) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.spacing = unit(0,'lines'))+
  theme( legend.position = "bottom" ) +
  labs(fill="Relative\nrisk")


``` 
